Course Project: Practical Machine Learning
========================================================
**Background Information:** *Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).*

**Goal of the project:** *Build machine learning algorithm to predict the manner in which they did the exercise. (This is the "classe" variable in the training set)*

```{r setoptions & load libraries}
#Just to make sure!
opts_chunk$set(echo = TRUE) 

library(caret)
library(randomForest)
```

### Data Processing:

Now, download the data if we haven't done so.
```{r download data if needed}
if (!file.exists("CourseProject")){
  dir.create("CourseProject")
  
  url_training = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
  url_testing = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
  
  download.file(url_training, destfile="CourseProject/training.csv", method="curl")
  download.file(url_testing, destfile="CourseProject/testing.csv", method="curl")}
```

Next, let's load the data into R.
```{r load data}
training = read.csv("CourseProject/training.csv", na.string=c("NA",""," "), as.is=TRUE)
testing = read.csv("CourseProject/training.csv", na.string=c("NA",""," "), as.is=TRUE)
```

Next, let's clean up the data & do some feature selection. In order to make sensible prediction of the outcome variable "classe", we 1) removed columns with NAs and 2) removed "metadata" that seemed to carry information about the data (e.g. user_name, timestamp etc), but w/o much sensible prediction power.
```{r clean and select data}
# 1) find columns with at least 1 NA value in its cells, and remove them
na_cols = names(training)[apply(training, 2, FUN=function(x) {any(is.na(x))})]
clean_training = training[, !(names(training) %in% na_cols)]
clean_testing = testing[,!(names(testing) %in% na_cols)]

# 2) remove extra uninteresting "metadata" columns
clean_training = clean_training[,8:length(names(clean_training))]
clean_testing = clean_testing[,8:length(names(clean_testing))]

clean_training$classe = as.factor(clean_training$classe)
clean_testing$classe = as.factor(clean_testing$classe)
```

### Data Analysis:
Split the clean training dataset & create a separate validation dataset for cross-validation.
```{r create cv}
set.seed(12345)
inTraining = createDataPartition(clean_training$classe, p=3/4, list=F)
final_training = clean_training[inTraining,]
final_validation = clean_training[-inTraining, ]
```

Select random forest as the machine learning algorithm to try first, as it has been suggested to be one of the top ML algorithm with high predictive power in contests.
```{r build random forest}
#fit_rf=train(as.factor(classe)~., data=final_training, method="rf")
# caret is painfully slow to fit random forest model.. use randomForest package instead
fit_rf= randomForest(as.factor(classe)~., data=final_training)
pred_training = predict(fit_rf, newdata=final_training)
confusionMatrix(pred_training, final_training$classe)
```
This model is working pretty well already, with accuracy of `r confusionMatrix(pred_training, final_training$classe)$overall[1]*100`% for in-sample prediction. (Also, the OOB (out-of-bag) estimate of error rate of this random forest model is 0.46%, pretty robust enough to continue with model cross-validation!). 

Now, time for some cross-validation Let's apply this working model to the validation dataset that we left behind earlier.
```{r run cross-validataion}
pred_cv = predict(fit_rf, newdata=final_validation)
confusionMatrix(pred_cv, final_validation$classe)
```
Still looking great! Cross-validation accuracy is at `r confusionMatrix(pred_cv, final_validation$classe)$overall[1]*100`%.

Lastly, let's try this model on a separate testing dataset to see how it performs out-of sample.
```{r run testing}
pred_testing = predict(fit_rf, newdata=clean_testing)
confusionMatrix(pred_testing, clean_testing$classe)
```
Super! We achieved an accuracy of `r confusionMatrix(pred_testing, clean_testing$classe)$overall[1]*100`%, even on a out-of-sample dataset. 

### Conclusion:
We were able to predict activities from measurement highly accurately with the random forest machine learning algorithm. 

